---
title: "p8105_hw2_jl5934"
author: "Jesus Luevano"
date: "2023-09-26"
output: github_document
---

```{r library}
library(tidyverse)
```

# Problem 1

##Importing the Data, cleaning it

Start with the polls data by month and year.

```{r pol Month}
Pol_Month.df = read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(
    mon,
    c("year", "month", "day")
  ) 

Pol_Month.df = Pol_Month.df %>% mutate(month = case_match(
    month,
    "01" ~ "January",
    "02" ~ "February",
    "03" ~ "March",
    "04" ~ "April",
    "05" ~ "May",
    "06" ~ "June",
    "07" ~ "July",
    "08" ~ "August",
    "09" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"))

Pol_Month.df = Pol_Month.df %>% mutate(President = case_when(
  prez_gop == 1 ~ "gop",
  prez_dem == 1 ~ "dem"
))
  
Pol_Month.df = select(Pol_Month.df, -c("prez_gop", "prez_dem", "day"))

```

Next do the Snp outcome, again by month and year. Here, the years were in the format of only the last two digits (as opposed to four digits like the other datasets). I saw that 00-13 related to 2000-2013, and 50-99 for 1950-1999. So I added a new column to create a prefix that I then concatenated to make the full four digit variable, and then removed that prior working column. This was to allow for downstream merging by year and month.

```{r snp}
Snp.df = read_csv("data/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>%
  separate(
    date,
    c("month", "day", "year")
  )

Snp.df = Snp.df %>% mutate(month = case_match(
    month,
    "1" ~ "January",
    "2" ~ "February",
    "3" ~ "March",
    "4" ~ "April",
    "5" ~ "May",
    "6" ~ "June",
    "7" ~ "July",
    "8" ~ "August",
    "9" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"))

Snp.df = select(Snp.df, c(year, month, everything()))

Snp.df <- Snp.df %>%
  mutate(year_prefix = case_when(
    year < 16 ~ 20,
    year > 49 ~ 19)) %>%
  mutate(year = paste(year_prefix,year, sep = "")) %>%
  select(-c("year_prefix"))


```

Last the unemployment rate by month and year. Here month information was by three letter abbreviation, so converted that. Additionally, year was in format of number, other two were character format, so I converted to allow for downstream merging. 

```{r unemployment}
Unemployment.df = read_csv("data/fivethirtyeight_datasets/unemployment.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_rate"
  )
Unemployment.df[["year"]] <- as.character(Unemployment.df[["year"]])

Unemployment.df = Unemployment.df %>% mutate(month = case_match(
    month,
    "jan" ~ "January",
    "feb" ~ "February",
    "mar" ~ "March",
    "apr" ~ "April",
    "may" ~ "May",
    "jun" ~ "June",
    "jul" ~ "July",
    "aug" ~ "August",
    "sep" ~ "September",
    "oct" ~ "October",
    "nov" ~ "November",
    "dec" ~ "December"))
```

## Merging the datasets

First we merge Pol data by month and snapshot based on same column of year and month. Then we take that new merged dataframe and merge it with unemployment again by year and month.

```{r merge datasets}
Pol_Snp.df <- full_join(Pol_Month.df, Snp.df, by = c("year", "month"))

Pol_Snp_Unemployment.df <- full_join(Pol_Snp.df, Unemployment.df, by = c("year", "month"))
```

## Describing the dataframes

Variables contained:

* The Pol_Month dataframe included variables for `r names(Pol_Month.df)`. The size of the dataset is `r nrow(Pol_Month.df)` rows by `r ncol(Pol_Month.df)` columns. It was describing the breakdown of seats in the senate, governorships, and the President based on Democrat or Republican status for specific date that was made up of a day, month, and year.

  * The range of years represented in this dataset are `r range(Pol_Month.df[["year"]])`.

* The Snp dataframe included variables `r names(Snp.df)`. The size of the dataset is `r nrow(Snp.df)` rows by `r ncol(Snp.df)` columns. It was describing the Snp close value, based on date that included day, month, and year.

  * The range of years represented in this dataset are `r range(Snp.df[["year"]])`. The range of the Snp was `r range(Snp.df[["close"]])`, and the mean was `r mean(Snp.df[["close"]])`.

* The Unemployment dataframe included variables `r names(Unemployment.df)`. The size of the dataset is `r nrow(Unemployment.df)` rows by `r ncol(Unemployment.df)` columns. It was describing the unemployment rate based on date that inclued month and year.

  * The range of years represented in this dataset are `r range(Unemployment.df[["year"]])`. When we attempt to caculate the range of the Unemployment rate `r range(Unemployment.df[["unemployment_rate"]])`, and the mean `r mean(Unemployment.df[["unemployment_rate"]])`, we get NA, as there are missing values that would need to be dealt with. 
  
  

# Problem 2

## Reading in and cleaning the Mr. Trash Wheel dataset

First, we will read in the Mr. Trash Wheel sheet from the xlsx file, selecing cells with data, ignoring the header icon and the sums at the bottom. Clean the column names. Then mutate a new variable for homes powered (based on the fact that each ton of trash equates to on average 500 kilowatts of electricity.  An average household will use 30). In this one year was a character, so it was chnaged to a numeric variable to allow for downstream merge. A new variable was made to label these values as belonging to Mr. Trash Wheel, labeled "Mister. 

```{r Mr. Trash Wheel}
MrTrashWheel.df <- readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx",
                   sheet = "Mr. Trash Wheel",
                   range = "A2:N549",
                   col_names = TRUE,
                   na = "") %>%
  janitor::clean_names()

MrTrashWheel.df <- MrTrashWheel.df %>% mutate(homes_powered = (weight_tons*500)/30)

MrTrashWheel.df <- MrTrashWheel.df %>% mutate(Trash_wheel = "Mister")

MrTrashWheel.df[["year"]] <- as.numeric(MrTrashWheel.df[["year"]])
```

Second, we will read in the Professor Trash Wheel sheet, selecting cells with data. Again we ignore the header and sums at the bottom. Column names are cleaned, and a new mutated variable for homes powered was made as above. A new variable was mutated to identify the source of the data, as "Professor".

```{r Professor Trash Wheel}
ProfTrashWheel.df <- readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx",
                   sheet = "Professor Trash Wheel",
                   range = "A2:M96",
                   col_names = TRUE,
                   na = "") %>%
  janitor::clean_names()

ProfTrashWheel.df <- ProfTrashWheel.df %>% mutate(homes_powered = (weight_tons*500)/30)

ProfTrashWheel.df <- ProfTrashWheel.df %>% mutate(Trash_wheel = "Professor")
```

Lastly, we will read in the Gwynnda Trash Wheel sheet. We selected cells with data only, ignoring the header and sums at the bottom. Column names were cleaned, and a new mutated variable for homes powered was made. Again a new variable was mutated to identify the source as "Gwynnda". 

```{r Gwynnda}
Gwynnda.df <- readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx",
                   sheet = "Gwynnda Trash Wheel",
                   range = "A2:K108",
                   col_names = TRUE,
                   na = "") %>%
  janitor::clean_names()

Gwynnda.df <- Gwynnda.df %>% mutate(homes_powered = (weight_tons*500)/30)

Gwynnda.df <- Gwynnda.df %>% mutate(Trash_wheel = "Gwynnda")
```

## Combine datasets

To combine datasets and retain all the information we first do full join with Mister and Professor Trash Wheel. Of note Professor Trash wheel has missing values for sports balls as it did not have that column to begin with. 

We then take that Mr_Prof set and combine it with Gwynnda, which of note adds a new variable column for plastic bags, but has missing values for glass bottles, grocery bags, chip bags, and sports balls as it did not include that information originally. 

```{r}
Mr_ProfTrashWheel.df <- full_join(MrTrashWheel.df, ProfTrashWheel.df) 

All_TrashWheel.df <- full_join(Mr_ProfTrashWheel.df, Gwynnda.df)
```


## Describing the combined dataset

* The variables contained in this combined dataset are `r names(All_TrashWheel.df)`. 

* The dimensions are `r ncol(All_TrashWheel.df)` columns by `r nrow(All_TrashWheel.df)` rows.

* The total weight of trash collected by professor Trash wheel was `r sum(ProfTrashWheel.df[["weight_tons"]])` tons. 

* The total number of cigarette butts collected by Gwynnda in July 2021 was `r sum(Gwynnda.df$cigarette_butts[which(Gwynnda.df$month == "July" & Gwynnda.df$year == '2021')])`. 


# Problem 3
