---
title: "p8105_hw2_jl5934"
author: "Jesus Luevano"
date: "2023-09-26"
output: github_document
---

```{r library}
library(tidyverse)
```

# Problem 1

##Importing the Data, cleaning it

Start with the polls data by month and year.

```{r pol Month}
Pol_Month.df = read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(
    mon,
    c("year", "month", "day")
  ) 

Pol_Month.df = Pol_Month.df %>% mutate(month = case_match(
    month,
    "01" ~ "January",
    "02" ~ "February",
    "03" ~ "March",
    "04" ~ "April",
    "05" ~ "May",
    "06" ~ "June",
    "07" ~ "July",
    "08" ~ "August",
    "09" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"))

Pol_Month.df = Pol_Month.df %>% mutate(President = case_when(
  prez_gop == 1 ~ "gop",
  prez_dem == 1 ~ "dem",
  prez_gop == 2 ~ "gop"
))
  
Pol_Month.df = select(Pol_Month.df, -c("prez_gop", "prez_dem", "day"))

```

Next do the Snp outcome, again by month and year. Here, the years were in the format of only the last two digits (as opposed to four digits like the other datasets). I saw that 00-13 related to 2000-2013, and 50-99 for 1950-1999. So I added a new column to create a prefix that I then concatenated to make the full four digit variable, and then removed that prior working column. This was to allow for downstream merging by year and month.

```{r snp}
Snp.df = read_csv("data/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>%
  separate(
    date,
    c("month", "day", "year")
  )

Snp.df = Snp.df %>% mutate(month = case_match(
    month,
    "1" ~ "January",
    "2" ~ "February",
    "3" ~ "March",
    "4" ~ "April",
    "5" ~ "May",
    "6" ~ "June",
    "7" ~ "July",
    "8" ~ "August",
    "9" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"))

Snp.df = select(Snp.df, c(year, month, everything()))

Snp.df <- Snp.df %>%
  mutate(year_prefix = case_when(
    year < 16 ~ 20,
    year > 49 ~ 19)) %>%
  mutate(year = paste(year_prefix,year, sep = "")) %>%
  select(-c("year_prefix", "day"))


```

Last the unemployment rate by month and year. Here month information was by three letter abbreviation, so converted that. Additionally, year was in format of number, other two were character format, so I converted to allow for downstream merging. 

```{r unemployment}
Unemployment.df = read_csv("data/fivethirtyeight_datasets/unemployment.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_rate"
  )
Unemployment.df[["year"]] <- as.character(Unemployment.df[["year"]])

Unemployment.df = Unemployment.df %>% mutate(month = case_match(
    month,
    "jan" ~ "January",
    "feb" ~ "February",
    "mar" ~ "March",
    "apr" ~ "April",
    "may" ~ "May",
    "jun" ~ "June",
    "jul" ~ "July",
    "aug" ~ "August",
    "sep" ~ "September",
    "oct" ~ "October",
    "nov" ~ "November",
    "dec" ~ "December"))
```

## Merging the datasets

First we merge Pol data by month and snapshot based on same column of year and month. Then we take that new merged dataframe and merge it with unemployment again by year and month.

```{r merge datasets}
Pol_Snp.df <- full_join(Pol_Month.df, Snp.df, by = c("year", "month"))

Pol_Snp_Unemployment.df <- full_join(Pol_Snp.df, Unemployment.df, by = c("year", "month"))


```

## Describing the dataframes

Variables contained:

* The Pol_Month dataframe included variables for `r names(Pol_Month.df)`. The size of the dataset is `r nrow(Pol_Month.df)` rows by `r ncol(Pol_Month.df)` columns. It was describing the breakdown of seats in the senate, governorships, and the President based on Democrat or Republican status for specific date that was made up of a day, month, and year.

  * The range of years represented in this dataset are `r range(Pol_Month.df[["year"]])`.

* The Snp dataframe included variables `r names(Snp.df)`. The size of the dataset is `r nrow(Snp.df)` rows by `r ncol(Snp.df)` columns. It was describing the Snp close value, based on date that included day, month, and year.

  * The range of years represented in this dataset are `r range(Snp.df[["year"]])`. The range of the Snp was `r range(Snp.df[["close"]])`, and the mean was `r mean(Snp.df[["close"]])`.

* The Unemployment dataframe included variables `r names(Unemployment.df)`. The size of the dataset is `r nrow(Unemployment.df)` rows by `r ncol(Unemployment.df)` columns. It was describing the unemployment rate based on date that inclued month and year.

  * The range of years represented in this dataset are `r range(Unemployment.df[["year"]])`. When we attempt to caculate the range of the Unemployment rate `r range(Unemployment.df[["unemployment_rate"]])`, and the mean `r mean(Unemployment.df[["unemployment_rate"]])`, we get NA, as there are missing values that would need to be dealt with. 
  
  

# Problem 2

## Reading in and cleaning the Mr. Trash Wheel dataset

First, we will read in the "Mr. Trash Wheel" sheet from the xlsx file, selecting cells with data, ignoring the header icon and the sums at the bottom. Next we clean the column names using the janitor package's function. Then we mutate a new variable for `homes_powered` (based on the fact that each ton of trash equates to on average 500 kilowatts of electricity,  An average household will use 30 tons). In this dataframe it had `year` as a character, so it was changed to a numeric variable to facilitate downstream merging. A new variable was made to label these values as belonging to Mr. Trash Wheel, titled `Trash_wheel == Mister`. 

```{r Mr. Trash Wheel}
MrTrashWheel.df <- readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx",
                   sheet = "Mr. Trash Wheel",
                   range = "A2:N549",
                   col_names = TRUE,
                   na = "") %>%
  janitor::clean_names()

MrTrashWheel.df <- MrTrashWheel.df %>% mutate(homes_powered = (weight_tons*500)/30)

MrTrashWheel.df <- MrTrashWheel.df %>% mutate(Trash_wheel = "Mister")

MrTrashWheel.df[["year"]] <- as.numeric(MrTrashWheel.df[["year"]])
```

Second, we will read in the "Professor Trash Wheel" sheet, selecting cells with data. Again we ignore the header and sums at the bottom. Column names are cleaned, and a new mutated variable for `homes_powered` was made as above as well as for the source of the data, where `Trash_wheel == "Professor`.

```{r Professor Trash Wheel}
ProfTrashWheel.df <- readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx",
                   sheet = "Professor Trash Wheel",
                   range = "A2:M96",
                   col_names = TRUE,
                   na = "") %>%
  janitor::clean_names()

ProfTrashWheel.df <- ProfTrashWheel.df %>% mutate(homes_powered = (weight_tons*500)/30)

ProfTrashWheel.df <- ProfTrashWheel.df %>% mutate(Trash_wheel = "Professor")
```

Lastly, we will read in the "Gwynnda Trash Wheel" sheet. We selected cells with data only, ignoring the header and sums at the bottom. Column names were cleaned, and a new mutated variable for `homes_powered` was made. Again a new variable was created, `Trash_wheel == Gwynnda`. 

```{r Gwynnda}
Gwynnda.df <- readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx",
                   sheet = "Gwynnda Trash Wheel",
                   range = "A2:K108",
                   col_names = TRUE,
                   na = "") %>%
  janitor::clean_names()

Gwynnda.df <- Gwynnda.df %>% mutate(homes_powered = (weight_tons*500)/30)

Gwynnda.df <- Gwynnda.df %>% mutate(Trash_wheel = "Gwynnda")
```

## Combine datasets

To combine datasets, and retain all the information, we first do `full_join` with Mister and Professor Trash Wheel. Of note Professor Trash wheel has missing values (NA) for sports balls as it did not have that column to begin with. 

We then take that `Mr_ProfTrashWheel.df` and combine it with Gwynnda, which of note adds a new variable column for plastic bags, but has missing values for glass bottles, grocery bags, chip bags, and sports balls as it did not include that information originally. This creates the `All_TrashWheel.df` with all the combined data for the three trash wheels.

```{r}
Mr_ProfTrashWheel.df <- full_join(MrTrashWheel.df, ProfTrashWheel.df) 

All_TrashWheel.df <- full_join(Mr_ProfTrashWheel.df, Gwynnda.df) 
```


## Describing the combined dataset

* The variables contained in this combined dataset are `r names(All_TrashWheel.df)`. 

* The dimensions are `r ncol(All_TrashWheel.df)` columns by `r nrow(All_TrashWheel.df)` rows.

* The total weight of trash collected by Professor Trash wheel was `r sum(ProfTrashWheel.df[["weight_tons"]])` tons. 

* The total number of cigarette butts collected by Gwynnda in July 2021 was `r sum(Gwynnda.df[["cigarette_butts"]][which(Gwynnda.df[["month"]] == "July" & Gwynnda.df[["year"]] == '2021')])`. 


# Problem 3

## Importing, cleaning, tidying up the dataset

We will Import the "MCI_Baseline.csv" file and create the `Baseline_demographics.df` table. Column names were cleaned using the `clean_names` function. 

* Next we wanted to change the format of the `sex` and `apoe4` variables, and made them both into binary factors, i.e. factors with two levels.

* Lastly, to subset out those that did develop MCI, we removed those with missing values of NA in the `age_of_onset` variable, and made a new table `Baseline_demographics_without_MCI.df`.

```{r baseline demographics}
Baseline_demographics.df <- read_csv("data/data_mci/MCI_baseline.csv", 
                                     skip = 1,
                                     na = ".") %>%
  janitor::clean_names()

Baseline_demographics.df[["sex"]] <- as.factor(Baseline_demographics.df[["sex"]]) #sex =1 if male, 0 = female, consider renaming
Baseline_demographics.df[["apoe4"]] <- as.factor(Baseline_demographics.df[["apoe4"]])

Baseline_demographics_without_MCI.df <- subset(Baseline_demographics.df, is.na(age_at_onset))

```

* A total of `r nrow(Baseline_demographics.df)` patients were recruited, but of those `r nrow(Baseline_demographics.df) - nrow(Baseline_demographics_without_MCI.df)` developed Mild Cognitive Impairment (MCI). When removed, we were left with `r nrow(Baseline_demographics_without_MCI.df)` patients in the study.

* Baseline mean age of all recruited participants was `r mean(Baseline_demographics.df[["current_age"]])`, and for those without MCI it was `r mean(Baseline_demographics_without_MCI.df[["current_age"]])`. Range of the ages for all recruited participants was `r range(Baseline_demographics.df[["current_age"]])`, and range of ages for those without MCI was `r range(Baseline_demographics_without_MCI.df[["current_age"]])`.

* The proportion of women in the study who are APOE4 carriers are `r (nrow(Baseline_demographics.df %>% filter(sex == '0' & apoe4 == '1')))/(nrow(Baseline_demographics.df %>% filter(sex == '0')))` from the total population, and `r (nrow(Baseline_demographics_without_MCI.df %>% filter(sex == '0' & apoe4 == '1')))/(nrow(Baseline_demographics_without_MCI.df %>% filter(sex == '0')))` among those without MCI.


### We will now import, clean, and rename the variable Study_ID to id to match the prior above in preparation for merging. 

```{r amyloid biomarker}
Amyloid_Biomarker.df <- read_csv("data/data_mci/mci_amyloid.csv",
                                 skip = 1,
                                 na = "NA") %>%
  janitor::clean_names() %>%
  rename("id" = "study_id")
```

We will determine which sample IDs are found in `Baseline_demographics_without_MCI.df` but not in `Amyloid_Biomarker.df`, and then conversely see which are in `Amyloid_Biomarker.df` but not in `Baseline_demographics_without_MCI.df`.

```{r see if shared samples}
Baseline_demographics_without_MCI.df[["shared_with_biomarkers"]] <- Baseline_demographics_without_MCI.df[["id"]] %in% c(Amyloid_Biomarker.df[["id"]])

Amyloid_Biomarker.df[["shared_with_baseline"]] <- Amyloid_Biomarker.df[["id"]] %in% c(Baseline_demographics_without_MCI.df[["id"]])
```

There are `r nrow(Baseline_demographics_without_MCI.df %>% filter(shared_with_biomarkers == "FALSE"))` IDs from the `Baseline_demographics_without_MCI.df` not found in the `Amyloid_biomaker` (id = 92, 179, 304, 389, 412). 

There are `r nrow(Amyloid_Biomarker.df %>% filter(shared_with_baseline == "FALSE"))` IDs from the `Amyloid_biomarker` not found in the `Baseline_demographics_without_MCI.df`.

## Combining the Datasets

We will use the `inner_join` function to combine the two tables of `Baseline_demographics_without_MCI.df` and `Amyloid_Biomarker.df`, so that only those with participants found in both are retained. 

* The table will then be saved as a CSV file to our local directory as "Baseline_amyloid.csv".

```{r combine if matching sample ID}
Baseline_Amyloid.df <- inner_join(Baseline_demographics_without_MCI.df, Amyloid_Biomarker.df, by = "id") %>%
  select(-c("shared_with_biomarkers", "shared_with_baseline"))

write_csv(Baseline_Amyloid.df, "data/data_mci/Baseline_amyloid.csv")
```

When looking at this new combined dataset it takes data based on `ID` and now includes both demographic information as well as clinical biomarkers, but only when a matching `ID` is found in both. 

* The variables included are `r names(Baseline_Amyloid.df)`.

* The size of the table is `r nrow(Baseline_Amyloid.df)` rows by `r ncol(Baseline_Amyloid.df)` columns. 


